---
title: "R for Marine Science Workshop 2 - Final Excercise"
author: "Miroslava Guerrero"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
 html_document:
    code_folding: show
    collapse: no
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: spacelab
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
# Set options, housekeeping
knitr::opts_chunk$set(
	echo = TRUE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)
```

# Keystone exercise: Estuary fish survey - Data rescue mission.

Messy data will be wrangled and final tables and figures will be produced. The data contains information on water quality, and predatory fish assemblages along the Ross River Estuary. The data contains information on 4 different sites and 4 species which was collected on a 30 day period. Along with metadata on the sites and environmental conditions from the sensor data.

## Housekeeping and libraries

```{r}
rm(list=ls()) # clear out environment

library(tidyverse)
library(readxl)
```

## Load main dataset

The catch data is split across 4 tabs in one excel file. Load them individually and then combine them together.

```{r}
catch1 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 1)
catch2 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 2)
catch3 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 3)
catch4 <- read_excel("../data/workshop2/estuary_catch_log.xlsx", sheet = 4)
```

Inspect each tab to look for abnormalities between workbooks and variable naming issues

```{r}
catch1
catch2
catch3
catch4
```

Although it does not seem to have any naming issues, it does. We need to combine the data and then check again

### Bind rows together

```{r}
catch_data <- bind_rows(catch1, catch2, catch3, catch4)
rm(catch1, catch2, catch3, catch4) # Clean up the environment
```

### Preliminary checks

```{r}
catch_data
head(catch_data)
glimpse(catch_data)
str(catch_data)
```

To properly check how many levels of site and species there are, we can do this by checking them by using distinct(), or converting them to factors. Firstly we will look at the distinct values first, in case we need to clean them before converting them to a factor.

```{r}
catch_data |> 
  distinct(site)

catch_data |> 
  distinct(species)
```

After running the above code we can observe we have 4 sites which seem to be different locations on the Ross River, corresponding to selected zones on the estuary. This will need to be checked against the metadata once loaded. Before the data gets joined with the metadata and taxonomic dictionary, we need to fix and clean the site and species names.

### Next steps for catch data

We need to fix the site names and place them in a reasonable order. As well as fix the species names by making them consistent and then using the taxonomic dictionary to translate to their scientific names.

## Load the metadata

```{r}
metadata <- read.csv("../data/workshop2/estuary_metadata.csv")
```

### Checks

```{r}
head(metadata)
glimpse(metadata)
```

The site names on the metadata are also all over the place, these will need to be standardized along with the catch_data table before both of the tables get combined.

### Fixing site names

The following code will fix the site names and convert any character variables to factor variables.

```{r}
# Fix site name and variables -data
catch_data <- catch_data |>
  mutate(
    site = str_to_lower(str_trim(site)), # Converts all letters to lowercase and removes extra spaces
    site = str_replace_all(site, " ", "_")) |> # replaces spaces with underscores
  mutate(across(where(is.character), as_factor) # Convert all chr to fct
  )

# Fix site name and variables - metadata
metadata <- metadata |>
  mutate(
    site_name = str_to_lower(str_trim(site_name)),
    site_name = str_replace_all(site_name, " ", "_")) |> 
  mutate(across(where(is.character), as_factor) # Convert all chr to fct
  )
```

### Recheck

```{r recheck-site-names}
# Check data table
str(catch_data)
# Check site names only
catch_data |> 
  distinct(site)

# Check metadata table
str(metadata)
# Check site names only
metadata |> 
  distinct(site_name)

```

Now the site names are fixed in both tables but there is still differences in the variable names (site vs site_name). This can be fixed when we join the datasets together.

### Fixing the species names

Using the taxonomic dictionary we can find the scientific names for the species. We need to match the common names in the catch log to match them to their scientific name before joining both datasets.

#### Loading the taxonomic dictionary

```{r}
species_dict <- read_csv("../data/workshop2/species_dictionary.csv")

# Check
head(species_dict)
```

After loading the taxonomic dictionary we can check the species name of both our data and the dictionary.

```{r}
# Species names we have
catch_data |> 
  distinct(species)

# Species name from dictionary
species_dict |>
  distinct(common_name)

```

As we can see they both have different names for the species. The common species name variable in the catch data to match those of the dictionary. Here we will fix the species names.

```{r}
catch_data <- catch_data |>
  mutate(
    species = str_to_lower(str_trim(species)),
    species = str_replace_all(species, " ", "_")
  )

catch_data |> 
  distinct(species)

```

Now we will create a new variable in the catch data by joining it with the taxonomic dictionary to indicate the scientific name.

```{r}
catch_data <- catch_data |>
  mutate(common_name = species) |>
  left_join(species_dict, by = join_by(common_name))

# Check that it worked
catch_data |> 
  distinct(species, scientific_name)

```

## Load the sensor data

The code for this was written in a different RScript which we can use to source.

The sourced script now contains daily means for three environmental variables (temperature, salinity, and turbidity) for each site, and any outliers were also removed.

```{r}
source(here::here("code/W2_SensorData_FinalEx.R"))

```

## Join catch data with the metadata and sensor data

```{r}
#Every row in catch_data is kept when using left_join()
catch_data_final <- catch_data |>
  left_join(metadata, by = join_by(site == site_name)) |> # This matches the site columns in both datasets
  left_join(sensor_daily, by = join_by(site, date)) # Matches rows where both the site and date are the same 

# Check
catch_data_final
glimpse(catch_data_final)
```

After checking we can see that we now have 427 rows of data when in reality we should have 480 \[4(spp.)\* 4(sites)\* 30(surveys) = 480\]. This happened as in some days there were 0 counts of certain species at different sites.

## Wrangle final dataset

We'll correct the missing rows by changing any missing data with 0's, and obtain the complete dataset with 480 rows. This will be important for any subsequent analyses and visualizations, as we want to be able to distinguish between true absences and missing data.

```{r}
catch_data_final <- 
catch_data_final |> 
  complete(
    nesting(site, date, 
            zone, mean_temp, mean_salinity, mean_turbidity),
    scientific_name
  ) |>
  mutate(
    count = coalesce(count, 0) # Convert the resulting NA counts to 0
  )

catch_data_final

```

After doing this we now have the 480 rows and we have finished our main data wrangling. Now we can proceed with any subsequent analyses to test hypotheses.

## Testing out hypotheses

### Difference in environmental variables

Firstly we will check if there is an actual difference in environmental variables among zones.

#### Salinity

```{r}
# Creating a summary table
sal_sum <- 
  catch_data_final |>
  group_by(zone) |>
  summarise(
    sd_salinity = sd(mean_salinity, na.rm = TRUE),
    median_salinity = median(mean_salinity, na.rm = TRUE),
    .groups = "drop"
  )

sal_sum

```

```{r}
# Create a plot from the summary

sal_sum |> 
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_salinity - sd_salinity, ymax = median_salinity + sd_salinity), width = 0.2) +
  theme_minimal() +
  labs(title = "Median Salinity by Zone (± SD)",
       x = "Zone",
       y = "Median Salinity (ppt)")

```

#### Turbidity

```{r}
# Create a summary table
turb_sum <- 
  catch_data_final |>
  group_by(zone) |>
  summarise(
    sd_turbidity = sd(mean_turbidity, na.rm = TRUE),
    median_turbidity = median(mean_turbidity, na.rm = TRUE),
    .groups = "drop"
  )

# Create a plot
turb_sum |>
  ggplot(aes(x = zone, y = median_turbidity)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_turbidity - sd_turbidity, ymax = median_turbidity + sd_turbidity), width = 0.2) +
  theme_minimal() +
  labs(title = "Median Turbidity by Zone (± SD)",
       x = "Zone",
       y = "Median Turbidity (NTU)")

```

#### Temperature

```{r}
temp_sum <- 
  catch_data_final |>
  group_by(zone) |>
  summarise(
    sd_temperature = sd(mean_temp, na.rm = TRUE),
    median_temperature = median(mean_temp, na.rm = TRUE),
    .groups = "drop"
  )

temp_sum |>
  ggplot(aes(x = zone, y = median_temperature)) +
  geom_point() +
  geom_errorbar(aes(ymin = median_temperature - sd_temperature, ymax = median_temperature + sd_temperature), width = 0.2) +
  theme_minimal() +
  labs(title = "Median Temperature by Zone (± SD)",
       x = "Zone",
       y = "Median Temperature (°C)")

```

Based on all three graphs we can deduce that only salinity differs between sites along the estuary. While temperature and turbidity vary very lightly between sites. 
We will therefore move on with salinity as our main environmental variation of interest, and discuss any other potentially confounding effects of temperature and turbidity in the discussion, as these could also potentially influence patterns of fish assemblages.

### Remaking the salinity plot

We will now remake the salinity plot so it will be publication ready.

```{r}
#| fig-cap: "Salinity across zones, Ross River Estuary System, Townsville, QLD (May 2026)."

sal_sum |> 
  ggplot(aes(x = zone, y = median_salinity)) +
  geom_point() +
  geom_errorbar(
    aes(
      ymin = median_salinity - sd_salinity,
      ymax = median_salinity + sd_salinity
    ),
    width = 0.2
  ) +
  theme_minimal() +
  labs(
    x = "Zone",
    y = "Median Salinity (ppt)"
  )

```

## Fish abundance patterns

This last graph aims to show the predatory fish abundance across the estuary salinity gradients in the Ross River Estuary system as of May 2026.

```{r}
# Summary stats
fish_stats <- catch_data_final |>
  group_by(zone, scientific_name) |>
  summarise(
    mean_catch = mean(count),
    sem_catch = sd(count) / sqrt(n()),
    .groups = "drop"
  )

fish_stats
```

```{r}
# Plot fish stats table

fish_stats |> 
  ggplot(aes(x = zone, y = mean_catch, fill = scientific_name)) +
  geom_line(group = 1) +
  geom_point(data = catch_data_final, aes(x = zone, y = count), # adds points from the final dataset
             position = position_jitter(width = 0.2), alpha = 0.2, size = 1, inherit.aes = FALSE) + 
  geom_point(position = "dodge", size = 4, pch = 23) +
  geom_errorbar(aes(ymin = mean_catch - sem_catch, ymax = mean_catch + sem_catch), # SE is narrow but allows viewer to see.
                position = position_dodge(0.9), width = 0.2) + 
  facet_wrap(~ scientific_name, scales = "fixed") +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "italic", size = 10, colour = "grey20"), # spp names in italic
    legend.position = "none" # legend unnecessary
  ) +
  labs(
    x = "Spatial Zone",
    # scientific label for y axis of n day -1 use regex to get superscript
    y = expression("Abundance (n day"^-1*")"),
  #  y = "Daily abundance (n day^-1)"
    y = "Abundance (per day)"
  )

```




